A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-09-02 bioc-workflow.Rmd
Finished tl section and modified pd section

```{r runPD, eval = FALSE}
## As per Christoforou et al (2016), 
pdRes <- phenoDisco(hl, fcol = "phenoDisco.Input", 
                    times = 200, 
                    GS = 60)

@@

```{r loadPDres, include=FALSE}
load("Data/pd_runs/pd.rda")
setStockcol(NULL)
setStockcol(paste0(getStockcol(), 70))
```
```{r showPDres}
pdRes
getMarkers(pdRes, fcol = "pd")
```

@@

```{r plotPDres, fig.width=9}
## Re-order the colours for the phenoDisco output
cl <- getMarkerClasses(pdRes, "pd")
cols <- getStockcol()[seq(cl)]
ind <- grep("Pheno", cl, invert = TRUE)
cols[ind] <- getStockcol()[seq(cl)][1:length(ind)]
cols[-ind] <- getStockcol()[seq(cl)][(length(ind) + 1):length(cl)]

## Plot the input and output
par(mfrow = c(1, 2))
plot2D(pdRes, fcol = "phenodisco.input", 
       main = "phenoDisco input markers", col = getStockcol()[1:6])
addLegend(pdRes, fcol = "phenodisco.input", 
          where = "bottomleft", cex = .7)
plot2D(pdRes, fcol = "pd", main = "phenoDisco output", col = cols)
addLegend(pdRes, fcol = "pd", where = "bottomleft", cex = .7, col = cols)
```

@@

```{r writePDres, eval=FALSE}
fData(pdRes)$pd <- as.character(fData(pdRes)$pd)
write.exprs(pdRes, fDataCols = "pd", file = "pd-results.csv", sep = ",")
```

@@

```{r visPDres, eval=FALSE}
pRolocVis(pdRes, fcol = "pd")
```

@@

```{r loadpdRes, include=FALSE}
pdRes <- hl 
load("Data/fusionparams.rda")
w <- table(getMarkers(pdRes, verbose = TRUE))
w <- 1/w[names(w) != "unknown"]
```

```{r optimise, eval=FALSE}
w <- table(getMarkers(pdRes, verbose = TRUE))
w <- 1/w[names(w) != "unknown"]

## 100 rounds of optimisation with five-fold cross-validation
params <- svmOptimisation(pdRes, fcol = "markers",
                          times = 100, xval = 5,
                          class.weights = w,
                          verbose = FALSE)
```

@@

```{r visualiseOpt, eval=FALSE}
(best <- getParams(params))
plot(params)
levelPlot(params)
```
```{r visualiseOptHide, echo=FALSE, fig.width = 12}
grid.arrange(plot(params), levelPlot(params), ncol = 2)
```

@@

```{r classify, eval=TRUE}
svmres <- svmClassification(pdRes, params,
                            class.weights = w,
                            fcol = "markers")
```

@@

```{r setColors, include=FALSE}
setStockcol(NULL)
setStockcol(paste0(getStockcol(), 70))
```
```{r plotSVM, eval=TRUE}
## set point size of each protein to be inversely proportional to the
ptsze <- exp(fData(svmres)$svm.scores) - 1

## plot new predictions
plot2D(svmres, fcol = "svm", cex = ptsze)
addLegend(svmres, fcol = "svm", where = "bottomleft", bty = "n", cex = .5)
```

The adjustment of the point size intuitively confers important
information that is more difficult to defined formally, but that we
will address in the next section. The classifier (SVM in our case, but
this is also valid of other classifiers) defines boundaries based on
the labelled marker proteins. These class/organelle boundaries define
how non-assigned proteins are classifier and with what
confidence. **TODO** explain how choosing markers
positively/negatively affects this, and relation with thresholding.

@@

It is common when applying a supervised classification algorithm to
set a specific score cutoff on which to define new assignments, below
which classifications are set to unknown/unassigned. This is important
as in a supervised learning setup proteins can only be predicted to be
localised to one of the sub-cellular niches that appear in the
labelled training data. We can not guaranette (and do not expect) that
the whole class diversity to be represented in the labelled training
data as (1) finding markers that represent the whole diversity of the
cell is challenging (especially obtaining dual- and multiply-localised
protein markers) and (2) many sub-cellular niches contain too few
proteins to train on (we recommend a minimum of 13 markers per
sub-cellular class for stratified 80/20 partitioning and 5-fold
cross-validation - this allows a minimum of ~10 examples for parameter
optimisation on the training partition i.e. 2 per fold for 5-fold
cross-validation, and then 3 for testing the best parameters on the
validation set). <!--Conversely, with defining markers one must also
be careful to not label too many markers, as this can lead to
overfitting and mis-assignment.-->

@@

```{r thresholds, fig.height=8, fig.width=8}
## First remove the markers
preds <- unknownMSnSet(svmres)

@@

```{r orgquants}
ts <- orgQuants(svmres, 
                fcol = "svm", 
                scol = "svm.scores",
                mcol = "markers", 
                t = .5)

preds <- getPredictions(svmres, 
                        fcol = "svm",
                        scol = "svm.scores", 
                        mcol = "markers",
                        t = ts)
```

@@

```{r plotres}
plot2D(preds, fcol = "svm.pred")
```

@@

```{r classifyApp, eval=FALSE}
library("pRolocGUI")
pRolocVis(svmres, app = "classify",
          fcol = "svm",
	        scol = "svm.scores",
          mcol = "markers")
```

@@

The first step, as with any machine learning algorithm, is to
optimise any free paramaters of the classifier. For the *k*-NN TL
classifier these are the class weights, one per subcellular niche,
that control the proportion of primary and auxiliary data to use for
learning. A weight can take any real value number between 0 and 1. A
weight of $\theta = 1$ indicates that all weight is given to the primary
data (and this implicitly implies that a weight of $1 - \theta$ is given to
the auxiliary data, similarly a weight of $\theta = 0$ implies that all
weight is given to the auxiliary data (so 0 to the primary source).
If we conduct a parameter seach and test weights $\theta = {0, 1/3, 2/3, 1}$
for each class, and if we have say 10 subcellular niches, this
will result in `4^10` different combinations of parameters to test. The
parameter optimisation is therefore time consuming and as such
we recommend you make use of a computing cluster.

<!--In the example below
we first examine a PCA plot of the data and overlay the remaining
subcellular markers. We can see that the two ribosomes, two nuclear
compartments and plasma membrane are quite well-defined and separated
form the rest of the plot, indicating that they are well-resolved in
the hyperLOPIT experiments. For this reason, and in the interest of
parameter optimisation and time we have removed these classes from the
$k$-NN TL optimisation as we expect their weight to be 1.

- add code here

We begin by performing the optimisation. The `knntlOptimisation`
function estimates the optimal weights for the different data sources
and the spatial niches described for the data.

- first need to do k-NN optimisation to get best k's for P and A
- explain weight matrix of thetas, `by` and `length,out`
- explain how to do this on the cluster, in practise too computationally heavy on a single processor

```{r tlopt, eval=FALSE}

opt <- knntlOptimisation(hl, gocc,
                         fcol = ,
                         by = .5, 
                         times = 1,
                         xval = 5, k = c(3,3),
                         BPPARAM = par
  )
```
-->

>>>>

```{r runPD, eval = FALSE}
## As per Christoforou et al (2016), 
hl <- phenoDisco(hl, fcol = "phenoDisco.Input", 
                 times = 200, 
                 GS = 60)

@@

```{r loadPDres, include=FALSE}
load("Data/pd_runs/pd.rda")
hl <- pdRes
setStockcol(NULL)
setStockcol(paste0(getStockcol(), 70))
```
```{r showPDres}
hl
getMarkers(hl, fcol = "pd")
```

@@

```{r plotPDres, fig.width=9}
## Re-order the colours for the phenoDisco output
cl <- getMarkerClasses(hl, "pd")
cols <- getStockcol()[seq(cl)]
ind <- grep("Pheno", cl, invert = TRUE)
cols[ind] <- getStockcol()[seq(cl)][1:length(ind)]
cols[-ind] <- getStockcol()[seq(cl)][(length(ind) + 1):length(cl)]

## Plot the input and output
par(mfrow = c(1, 2))
plot2D(hl, fcol = "phenodisco.input", 
       main = "phenoDisco input markers", col = getStockcol()[1:6])
addLegend(hl, fcol = "phenodisco.input", cex = .7)
plot2D(hl, fcol = "pd", main = "phenoDisco output", col = cols)
addLegend(hl, fcol = "pd", cex = .7, col = cols)
```

@@

```{r writehl, eval=FALSE}
fData(hl)$pd <- as.character(fData(hl)$pd)
write.exprs(hl, fDataCols = "pd", file = "pd-results.csv", sep = ",")
```

@@

```{r vishl, eval=FALSE}
pRolocVis(hl, fcol = "pd")
```

From examining the literature and the Gene Ontology we find that
phenotype 1 is enriched in proteins that localise to the nucleus
(non-chromatin associated proteins), phenotype 2 is enriched in
chromatin associated proteins, phenotype 3 contains many cytosolic
proteins and phenotypes 4 and 5 contain lysosmal and endosomal
proteins. From curating these phenotypes we obtain new marker proteins
that can be used as input for supervised classification.

@@

```{r loadpdRes, include=FALSE}
load("Data/fusionparams.rda")
w <- table(getMarkers(hl, verbose = TRUE))
w <- 1/w[names(w) != "unknown"]
```

```{r optimise, eval=FALSE}
w <- table(getMarkers(hl, verbose = TRUE))
w <- 1/w[names(w) != "unknown"]

## 100 rounds of optimisation with five-fold cross-validation
params <- svmOptimisation(hl, fcol = "markers",
                          times = 100, xval = 5,
                          class.weights = w)
```

@@

```{r visualiseOpt, eval=FALSE}
(best <- getParams(params))
plot(params)
levelPlot(params)
```
```{r visualiseOptHide, echo=FALSE, fig.width = 12}
library('gridExtra')
grid.arrange(plot(params), levelPlot(params), ncol = 2)
```

@@

```{r classify, eval=TRUE}
hl <- svmClassification(hl, params,
                        class.weights = w,
                        fcol = "markers")
```

@@

```{r setColors, include=FALSE}
setStockcol(NULL)
setStockcol(paste0(getStockcol(), 70))
par(mfrow = c(1, 1))
```
```{r plotSVM, eval=TRUE}
## set point size of each protein to be inversely proportional to the
ptsze <- exp(fData(hl)$svm.scores) - 1

## plot new predictions
plot2D(hl, fcol = "svm", cex = ptsze)
addLegend(hl, fcol = "svm", where = "bottomleft", bty = "n", cex = .5)
```

The adjustment of the point size intuitively confers important
information that is more difficult to defined formally, but that we
will address in the next section. The classifier (SVM in our case, but
this is also valid of other classifiers) defines boundaries based on
the labelled marker proteins. These class/organelle boundaries define
how non-assigned proteins are classified and with what
confidence. **TODO** explain how choosing markers
positively/negatively affects this, and relation with thresholding.

@@

It is common when applying a supervised classification algorithm to
set a specific score cutoff on which to define new assignments, below
which classifications are set to unknown/unassigned. This is important
as in a supervised learning setup proteins can only be predicted to be
localised to one of the sub-cellular niches that appear in the
labelled training data. We can not guaranette (and do not expect) that
the whole class diversity to be represented in the labelled training
data as (1) finding markers that represent the whole diversity of the
cell is challenging (especially obtaining dual- and multiply-localised
protein markers) and (2) many sub-cellular niches contain too few
proteins to train on (we recommend a minimum of 13 markers per
sub-cellular class for stratified 80/20 partitioning and 5-fold
cross-validation - this allows a minimum of ~10 examples for parameter
optimisation on the training partition i.e. 2 per fold for 5-fold
cross-validation, and then ~3 for testing the best parameters on the
validation set). <!--Conversely, when defining markers one must also
be careful to not label too many markers, as this can lead to
overfitting and mis-assignment.-->

@@

```{r thresholds, fig.height=8, fig.width=8}
## First remove the markers
preds <- unknownMSnSet(hl)

@@

```{r orgquants}
ts <- orgQuants(hl, 
                fcol = "svm", 
                scol = "svm.scores",
                mcol = "markers", 
                t = .5)

hl <- getPredictions(hl,
                     fcol = "svm",
                     scol = "svm.scores", 
                     mcol = "markers",
                     t = ts)

@@

```{r resetpar}
dev.off()
```
```{r plotres}
plot2D(hl, fcol = "svm.pred")
```

@@

```{r classifyApp, eval=FALSE}
library("pRolocGUI")
pRolocVis(hl, 
          app = "classify",
          fcol = "svm",
	        scol = "svm.scores",
          mcol = "markers")
```

@@

The first step, as with any machine learning algorithm, is to optimise
any free paramaters of the classifier. For the *k*-NN TL classifier
there are two sets of parameters that need optimising: the first set
are the $k$'s for the primary and auxiliary data sources required for
the nearest neighbour calculations for each data source.  The second
set of parameters (which is noted by a vector of $\theta$ weights)
that require optimising are the class weights, one per subcellular
niche, that control the proportion of primary and auxiliary data to
use for learning. A weight can take any real value number between 0
and 1. A weight of $\theta = 1$ indicates that all weight is given to
the primary data (and this implicitly implies that a weight of $1 -
\theta$ is given to the auxiliary data, similarly a weight of $\theta
= 0$ implies that all weight is given to the auxiliary data (so 0 is
given to the primary source). If we conduct a parameter seach and test
weights $\theta = {0, 1/3, 2/3, 1}$ for each class, and if we have,
for example 10 subcellular niches, this will result in `4^10`
different combinations of parameters to test. The parameter
optimisation is therefore time consuming and as such we recommend you
make use of a computing cluster (code and submissing scripts are also
available in the supporting information). The markers in the `hl`
dataset contain 14 subcellular classes. If we examine these markers
and classes on the PCA plot above we can see that in particular the
two ribosomes and two nuclear compartments are highly separated along
the first two components, this is also evident from the profiles plot
which gives us a good indication that these subcellular niches are
well-resolved in the hyperLOPIT dataset. Transfer learning is
particularly useful for classes that are not as well separated, we
find that subcellular niches that are well-separated under hyperLOPIT
and LOPIT obtain a class score of 1 (i.e. use only primary data from
transfer learning [@Breckels:2016]). Therefore, for the optimisation
stage of the analyses we can already infer a subcellular class weight
of 1 for these niches and only optimise over the remaining organelles.
This can signifciantly cut down optimisation time as by removing these
4 classes from the optimisation (and not the classification) we only
have `4^10` class weight combinations to consider instead of `4^14`
combinations. In the example below we remove these 4 classes from the
marker set, re-run the `knnOptimisation` for each data source and then
run the `knntlOptimisation` with the 10 remaining classes. (Note: this
is not run live as this the `hl` dataset with 10 classes, 707 markers
and `4^10` combinations of parameters takes ~76 hours to run on the
University of Cambridge HPC using 256 workers).

In the example code below we first remove the 2 ribosomes and nuclear
classes and then run the `knnOptimisation` followed by the
`knntlOptimisation` (specifying snow with a MPI
backend with 255 workers for parallelisation on the Cambridge HPC).

```{r tlopt, eval=FALSE}
## create new markers column for tl markers
fData(hl)$tlmarkers <- fData(hl)$markers
fData(gocc)$tlmarkers <- fData(gocc)$markers

## Remove 4 classes
torm  <- c("40S Ribosome", "60S Ribosome",
           "Nucleus - Chromatin", 
           "Nucleus - Non-chromatin")
for (i in seq(torm)) {
  hl <- fDataToUnknown(hl, from = torm[i], fcol = "tlmarkers")
  gocc <- fDataToUnknown(hl, from = torm[i], fcol = "tlmarkers")
}
getMarkerClasses(hl, fcol = "tlmarkers")
getMarkerClasses(hl, fcol = "tlmarkers")

## get best k's
kp <- knnOptimisation(hl, fcol = "tlmarkers")
ka <- knnOptimisation(gocc, fcol = "tlmarkers")

## Set appropriate backend and workers for the tl
par <- SnowParam(255L, type = "MPI")

## Now peform tl optimisation 
tlopt <- knntlOptimisation(hl, gocc,
                           fcol = "tlmarkers",
                           length.out = 4, 
                           times = 50,
                           xval = 5, k = c(kp,ka),
                           BPPARAM = par
  )
```
```{r loadtlres, include = FALSE}
load("Data/tl_runs/tlopt.rda")
```

The results of the optimisation can be visalised using the `plot`
method for `"ThetaRegRes"` objects:

```{r plottl}
plot(tlopt)
```

Looking at the bubble plot displaying the distribution of best weights
over the 50 runs we find that for many of the subcellular niches a
weight of 1 is most popular (i.e. use only primary hyperLOPIT data in
classification), this is unsuprising as we already know the dataset is
well resolved for these classes. We see that the most popular weights
for the proteasome and lysosome tend to be towards 0, indicating that
these niches are well-resolved in the Gene Ontology. This tells us
that we would benefit from including auxiliary GO information in our
classifier for these subcellular compartments. The plasma membrane
weights are relatively equally spread between using hyperLOPIT and GO
data. Using the `getParams` function we can return the best weights
and then use this as input for the classification.

One of the benefits of the algorithm is the ability to manually select
weights for each class. In the optimisation above, for time
constraints, we removed the two ribosomal subunits and the two nulcear
compartments, and therefore in the code chunk below when we extract
the best parameters, these subcellular niches are not included. To
include these 4 subcellular niches in the next classification step we
must include them in the parameters so we infer a weight of 1 for each
of these niches as we know they are well resolved in hyperLOPIT. We
then re-order the weights according to `getMarkerClasses` and perform
the classification using the function `knntlClassification`.

```{r tlclassify}
## best parameters for the 10 classes
(bestpar <- getParams(tlopt))

## add weights for classes not included in the optimisation
otherweights <- rep(1, 4)
names(otherweights) <- c("40S Ribosome", "60S Ribosome", 
                         "Nucleus - Chromatin", 
                         "Nucleus - Non-chromatin")
(bestpar <- c(bestpar, otherweights))

## re-order classes 
bestpar <- bestpar[getMarkerClasses(hl)]

## Do the classification
hl <- knntlClassification(hl, gocc, bestTheta = bestpar, 
                          k = c(kp, ka))

The results from the classification results and associated scores are
appended to the `fData` slot and named `knntl` and `knntl.scores`
respectively. Results can be visualised using `plot2D`, scores
assessed and cutoffs calculated using the `classify` app in
`pRolocVis`, predictions obtained using `getPredictions` in the same
way as demonstrated above for the SVM classifier.
